\chapter{Summary}
\index{Summary}

This report provides an in-depth view of my second internship at Delem. The result is a program, which I expect to be very useful for Delem. The intention was to create this program as extensible and maintainable as possible. This result has fully been accomplished.

The reader of this report is expected to be familiar with HSB modules. It is advisable to read the background information on page \pageref{achtergrondinfo} as well as the project description on page \pageref{project} if this knowlege is absent.

During this internship, I've spent about one day a week on other activities. A good example of this is testing an existing application VBend\footnote{Virtual Bend, application which is capable of determining and simulating an optimal bending sequencetesting an existing bendsequence}, which would soon be released. During this activity, I was fully integrated within the project team itself and this experience has given me a pretty good indication on 'normal' projects at Delem. The usuage of \index{PR}PR and \index{CR}CR\footnote{Problem Report/Change Request, refer to page \pageref{PR} for more information}'s was also greatly illustrated here.

I've also worked quite a bit with XML, which could prove useful in the future. Futhermore, my Windows programming wasn't quite optimal (I hardly use Windows at home), so I learned quite a bit there as well.

Delem has given me quite some freedom while carrying out this internship, which I attribute to the fact that I've been employed at Delem for over 2 years now. However, I believe this internship has given me the opportunity to learn the ins and outs of projectmanagement at Delem, as wel as the company itself.

Finally, I believe this has been a very good internship. The project has been well documented and the program is quite functional. It has already been used to find seveneral issues, which may be addressed in the future. A good result, which my supervisor at Delem confirmed.
